---
title: "CRISPR_9 analysis from SMB_21 cells from https://pmc.ncbi.nlm.nih.gov/articles/PMC11304869/
"
author: "Carlos Alfaro"
date: "2025-06-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r Libraries}
suppressPackageStartupMessages({
  library(scater);library(Seurat);library(tidyverse);library(cowplot);library(Matrix.utils);library(edgeR)
  ;library(dplyr);library(magrittr);library(Matrix);library(purrr);library(reshape2);library(S4Vectors)
  ;library(tibble);library(SingleCellExperiment);library(pheatmap);library(apeglm);library(png)
  ;library(DESeq2);library(RColorBrewer);library(clusterProfiler);library(org.Mm.eg.db);library(CellChat);
  library(patchwork);library(enrichplot);library(STRINGdb);library(org.Hs.eg.db); library(homologene)})
```


```{r Pre-processing}
# Loading CRISPR-9 counts 
SMB21_counts<- read.csv("data/CRISPR_screen_counts_SMB21.txt", sep = "\t")
# Loading matching reference for gene identification
# Merge count file (with sgRNA barcodes) and the Brie .chip file (which contains barcode → gene mappings), then save the annotated result.
brie_chip <- read_tsv("CP0044_GRCm38_NCBI_CRISPRko_strict_gene_20221201.chip")
# Rename columns to prepare for merge
brie_chip <- brie_chip %>%rename(`Construct.Barcode` = `Barcode Sequence`,Gene = `Gene Symbol`)
# Merge based on barcode
annotated_counts <- SMB21_counts %>% left_join(brie_chip, by = "Construct.Barcode")
#Reorder columns (optional, for clarity)
annotated_counts <- annotated_counts %>%select(`Construct.Barcode`, `Construct.IDs`, Gene, starts_with("REP"), plasmid)
# Write to file
write_tsv(annotated_counts, "annotated_counts_SMB21.tsv")
```

```{r DAOY counts}
DAOY_counts<- read.csv("CRISPR_sceen_counts_DAOY.txt", sep = "\t")
brunello_chip <- read_tsv("CP0043_GRCh38_NCBI_CRISPRko_strict_gene_20240130.chip")
#Rename columns to prepare for merge
brunello_chip <- brunello_chip %>%rename(`Construct.Barcode` = `Barcode Sequence`,Gene = `Gene Symbol`)
# Merge based on barcode
annotated_counts1 <- DAOY_counts %>%left_join(brunello_chip, by = "Construct.Barcode")
# Reorder columns (optional, for clarity)
annotated_counts1 <- annotated_counts1 %>%select(`Construct.Barcode`, `Construct.IDs`, Gene,starts_with("DAOY"))
# Write to file
write_tsv(annotated_counts1, "output/annotated_counts_DAOY.tsv")
```

```{r SMB21 human ortologs}
# Renaming the column names on the dataframe
colnames(annotated_counts)
annotated_counts<- annotated_counts |> dplyr::rename(sgRNA = Construct.Barcode)
annotated_counts<- annotated_counts |> dplyr::rename(Rep1 = REP.A.SHH.MB , Rep2 = REP.B.SHH.MB, Rep3 = REP.C.SHH.MB )
annotated_counts<- annotated_counts |> dplyr::rename(gene = Gene)
annotated_counts<- annotated_counts |> dplyr::select(-Construct.IDs)
# Saving annotated text 
annotated_counts[duplicated(annotated_counts$sgRNA),]
write.table(annotated_counts, "output/annotated_counts_SMB21.txt", sep = "\t", col.names= T, row.names = F)
# Fixing quotes
annotated <- read_tsv("annotated_counts_SMB21.txt")
# Write it again WITHOUT quotes
write_tsv(annotated, "output/mageck_ready_counts.txt", quote_escape = "none")
```

```{r Inspecting ouput }
output_gene<- read.csv("mouse_output.gene_summary.txt", sep = "\t")
head(output_gene)
# SAVING SIG GENES ( NEGATUVELY REGULATED)
write.csv(sig_genes, file ="Sign_genes.csv")
```

#### Processing of the ouput ########
###### Loading the non-core essential genes in the results from the cell line and do gene ontology and KEEG analysis #####
##########################################################################################################################
```{r}
# Sig genes will be the essential genes identified as the negatively selected 
#sig_genes<- output_gene |>
## Create tmp file that indicates the most negative selected genes and then do Gene ontology 
## Loading non-core genes 
non_core_genes <- read.csv("non_essential_mouse_genes.csv")
tmp2<- non_core_genes |> pull(id)
GO_analysis <- function(gene_list, condition, status) {
  base_path<-paste0("GO/", condition, "/", status)
  if (!dir.exists(base_path)) {dir.create(base_path, recursive = TRUE)}
  set.seed(0708)
  GO <- enrichGO(gene = gene_list,OrgDb = "org.Mm.eg.db",keyType = "SYMBOL",ont = "MF",
    pAdjustMethod = "BH",pvalueCutoff = 0.05, minGSSize = 10,  maxGSSize = 500 )
  
  write.csv(as.data.frame(GO), file = paste0(base_path, "/", status, ".MF.GO.csv"))
  pdf(file = paste0(base_path, "/", status, ".MF.GO.pdf"), width = 10, height = 5, useDingbats = FALSE)
  print(dotplot(GO, label_format = 100, showCategory = 20) +ggtitle(paste(status, condition, sep = " - ")) +
      theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold")))
  dev.off()
  
  enrich_sim <- pairwise_termsim(GO)
  pdf(file = paste0(base_path, "/", status, ".MF.Enrichment.pdf"), width = 9, height = 5, useDingbats = FALSE)
  print(emapplot(enrich_sim, showCategory = 20) +ggtitle(paste(status, "Enrichment map", condition, sep = " - ")) +
      theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold")))
  dev.off()
  rm(GO, enrich_sim)
  gc()}

## Running the finction 
GO_analysis(tmp2, "SMB21", "All")
```

# Extending the analysis of ubiquitin and check if SSH medulloblastom pathways are related with proteosome.
```{r}
tmp2<- genes |> pull(id)
set.seed(0708)
GO <- enrichGO(gene = tmp2,OrgDb = "org.Mm.eg.db",keyType = "SYMBOL",ont = "BP",
    pAdjustMethod = "BH",pvalueCutoff = 0.05,minGSSize = 10, maxGSSize = 500  )
GO@result[c(362,436),] # 362, 436, 450, 542
ups_terms <- GO@result %>% filter(grepl("proteasome|ubiquitin|proteasomal", Description, ignore.case = TRUE)) ###proteasome|ubiquitin|smoothened|cilium|proteasomal
ups_genes <- unlist(strsplit(paste(ups_terms$geneID, collapse = "/"), split = "/"))

## String 
string_db <- STRINGdb$new(version = "11.5", species = 10090, score_threshold = 400)
# Combine SHH and UPS genes
# Map to STRING IDs
mapped <- string_db$map(data.frame(gene = ups_genes), "gene", removeUnmappedRows = TRUE)
# Plot interaction network
pdf("string_tetsing.pdf", width = 10, height = 10)
string_db$plot_network(mapped$STRING_id)
invisible(dev.off())

## SHH PATHWAY
shh_genes_mouse <- c(
  "Shh",     # Sonic hedgehog
  "Ptch1",   # Patched 1
  "Ptch2",   # Patched 2
  "Smo",     # Smoothened
  "Gli1",    # GLI family zinc finger 1
  "Gli2",    # GLI family zinc finger 2
  "Gli3",    # GLI family zinc finger 3
  "Sufu",    # Suppressor of fused
  "Spop",    # Speckle-type POZ protein
  "Cul3",    # Cullin 3
  "Skp1a",   # S-phase kinase-associated protein 1A
  "Skp2",    # S-phase kinase-associated protein 2
  "Rbx1",    # Ring-box 1
  "Huwe1",   # HECT, UBA and WWE domain containing 1, E3 ubiquitin protein ligase
  "Mycn"     # N-Myc proto-oncogene
)

ups_terms$has_SHH_gene <- sapply(ups_terms$geneID, function(gene_list) {any(strsplit(gene_list, "/")[[1]] %in% shh_genes_mouse)})
ups_terms_with_shh <- ups_terms %>% filter(has_SHH_gene)
filtered_terms <- ups_terms %>% filter(has_SHH_gene == TRUE)  
# Subset the enrichment object (ego_mouse)
ego_filtered <- GO
ego_filtered@result <- GO@result %>%filter(ID %in% ups_terms$ID)
ego_filtered@result <- ego_filtered@result[!duplicated(ego_filtered@result$ID), ]  
enrich_sim <- pairwise_termsim(ego_filtered)
emapplot(enrich_sim, showCategory = 200)
dotplot(ego_filtered)
cnetplot(ego_filtered)
## Enrichment 
#enrich_sim <- pairwise_termsim(ups_terms)
pdf("tetsing_emmaplot1.pdf", width = 10, height = 10)
emapplot(enrich_sim, showCategory = 30) + theme(legend.position = "none")
invisible(dev.off())
```

```{r KEEG}
KEGG_analysis_bulk <- function(gene_list, condition, status, output_base = "KEGG/") {
  ## Create output directory
  base_path <- file.path(output_base, condition, status)
  if (!dir.exists(base_path)) {dir.create(base_path, recursive = TRUE)}

  set.seed(0708)
  # Convert gene symbols to Entrez IDs (human)
  entrez_ids <- bitr(gene_list, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Mm.eg.db)
  if (nrow(entrez_ids) == 0) {stop("❌ No valid Entrez IDs found for the input gene symbols.")}
  entrez_gene_list <- entrez_ids$ENTREZID
  message("✅ Total Entrez IDs mapped: ", nrow(entrez_ids))
  # Run KEGG enrichment analysis
  kegg_results <- enrichKEGG(gene = entrez_gene_list,organism = "mmu",minGSSize = 10,maxGSSize = 500,pvalueCutoff = 0.05)
  ##Save results
  write.csv(as.data.frame(kegg_results), file = file.path(base_path, paste0(status, "_KEGG.csv")))
  ## Dotplot
  pdf(file = file.path(base_path, paste0(status, "_KEGG_dotplot.pdf")), width = 9, height = 5, useDingbats = FALSE)
  print(dotplot(kegg_results, label_format = 100, showCategory = 20) +ggtitle(paste(status, condition, sep = " - ")) +
      theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold")))
  dev.off()
  ## Enrichment map
  enrich_sim <- pairwise_termsim(kegg_results)
  pdf(file = file.path(base_path, paste0(status, "_KEGG_enrichmentmap.pdf")), width = 9, height = 5, useDingbats = FALSE)
  print(emapplot(enrich_sim, showCategory = 20) +ggtitle(paste(status, "Enrichment map", condition, sep = " - ")) +
      theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold")))
  dev.off()
  ## Return result for downstream use
  return(kegg_results)}

kegg_res <- KEGG_analysis_bulk(gene_list = tmp2,condition = "SMB21",status = "All")
```

```{r KEEG in chosen pathways}
# Set seed for reproducibility
set.seed(0708)
# Convert gene symbols to Entrez IDs (human)
entrez_ids <- bitr(tmp2, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Mm.eg.db)
entrez_gene_list <- entrez_ids$ENTREZID

kegg_results <- enrichKEGG(gene = entrez_gene_list,organism = "mmu",  minGSSize = 10,maxGSSize = 500,pvalueCutoff = 0.05)

## Dotplot by fold enrichment ( GeneRatio/ BgRatio)
# It shows how much more frequent the pathway appears in your gene list vs. random expectation. 
dotplot(kegg_results, x = "FoldEnrichment", label_format = 100, showCategory = 20) +ggtitle("something") +  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))

dotplot(kegg_results, label_format = 100, showCategory = 20) + ggtitle("something") + theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))

## Picking significant pathways.
kegg_df <- kegg_results@result
selected_paths <- c("Ribosome", "Spliceosome",
                    "Cell cycle", "Oxidative phosphorylation",
                    "Ubiquitin mediated proteolysis", "Proteasome",
                    "Nucleotide excision repair","ATP-dependent chromatin remodeling",
                    "Neutrophil extracellular trap formation", "DNA replication",
                    "RNA degradation","mTOR signaling pathway",
                    "PI3K-Akt signaling pathway", "Cellular senescence",
                    "Base excision repair","Wnt signaling pathway",
                    "Protein export","Hedgehog signaling pathway",
                    "HIF-1 signaling pathway", "cAMP signaling pathway",
                    "Mismatch repair")
kegg_subset <- kegg_df %>% filter(Description %in% selected_paths)
# Make a copy and overwrite the result slot
filtered_kegg <- kegg_results
filtered_kegg@result <- kegg_subset
# Plot
pdf("KEGG/Maria_plots/Generatio_picked_pathways.pdf", width = 8, height = 6)
dotplot(filtered_kegg, showCategory = length(selected_paths), label_format = 40) +
  ggtitle("Input gene coverage") +theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
        axis.text.y = element_text(size = 14), axis.text.x = element_text(size = 14), 
        axis.title.x = element_text(size = 15), axis.title.y = element_text(size = 14))   
invisible(dev.off())

pdf("KEGG/Maria_plots/FoldEnrich_picked_pathways.pdf", width = 8, height = 6)
dotplot(filtered_kegg, x ="FoldEnrichment",showCategory = length(selected_paths),label_format = 40) +
  ggtitle("Enrichment Score") + theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
          axis.text.y = element_text(size = 14),axis.text.x = element_text(size = 15),
          axis.title.x = element_text(size = 15),axis.title.y = element_text(size = 14))    
invisible(dev.off())
```





###### Proteosome activity of the genes by comparing with Reactome#####
-----------------------------------------------------------------------

```{r}
### Loading the genes 
genes<- read.csv("non_essential_mouse_genes.csv", sep = ",")
#install.packages("msigdbr")
library(clusterProfiler)
library(org.Mm.eg.db)     # Use the mouse annotation database
library(msigdbr)
library(dplyr)

mageck_genes<- non_core_genes$id
# Step 1: Convert mouse gene symbols to Entrez IDs
gene_entrez <- bitr(mageck_genes, 
                    fromType = "SYMBOL", 
                    toType = "ENTREZID", 
                    OrgDb = org.Mm.eg.db)
# Load MSigDB mouse Reactome gene sets (category C2, subcategory CP:REACTOME)
msigdb_mouse <- msigdbr(species = "Mus musculus", category = "C2", subcategory = "CP:REACTOME")

# Step 3: Filter for proteasome-related pathways
msigdb_proteasome <- msigdb_mouse %>%
  filter(grepl("PROTEASOME", gs_name, ignore.case = TRUE))

# Step 4: Run enrichment
enrich_res <- enricher(gene = gene_entrez$ENTREZID,
                       TERM2GENE = msigdb_proteasome[, c("gs_name", "entrez_gene")])

# Optional: View results
head(enrich_res)
```

# Removing the core essential genes: Genes required for most cells for basic survival  
```{r Druggable targets}
pan_essentials<- read.csv("pan-species-control-essentials-50genes.txt", sep = "\t")
# Selecting genes that are non-essential 
sig_genes_non_essential<- sig_genes|>  dplyr::filter(!id %in% pan_essentials$Mouse) 
## Transforming the non-essential genes into human ortologs
genes<- sig_genes_non_essential |>pull(id)
write.csv(sig_genes_non_essential, file = "non_essential_mouse_genes.csv")
# Map to human orthologs
orthologs <- homologene(genes, inTax = 10090, outTax = 9606) 
human_genes <- unique(orthologs$`9606`)
genes_clean <- as.character(human_genes)
genes_clean <- genes_clean[!is.na(genes_clean) & genes_clean != ""]
### Loading interactions.tsv 
interatcions<- read.csv("interactions.tsv", header = T, sep = "\t")
```

```{r}
Interactions_genes<- genes_clean %in% interatcions$gene_name
results<-interatcions[Interactions_genes,]
length(unique(results$gene_name))
results1<- results|> filter(approved == "TRUE")
length(unique(results1$gene_name))
```

```{r r session-info}
sessioninfo::session_info()
```
